The First Level of Super Mario Bros.~is Easy with Lexicographic Orderings and Time Travel {\normalsize \ldots after that it gets a little tricky.}

### 2013-03-17 tasbot/paper/paper.tex
MARIONET section

\newcommand\comment[1]{}
\newcommand\parameteralert[1]{
  % XXX this can get cooler.
  \begin{quotation}
  \noindent {\bf Parameter Alert!}\, #1
  \end{quotation}
}

@@

\paragraph{Takes of woe, tales of joy.} 
Backtracking was initially added in order to fix the mario coin ledge
problem in {\tt 1-2}. It did not help with this. Older versions of the
code would have Mario get past this spot, probably by luck, but in
these he would get stuck in an embarrassing way on a small pit later,
or jump straight into an enemy. Most of the approaches that actually
looked solidly good elsewhere were failing badly here. As I started
writing up this paper, on an airplane, I discovered the bug in the
weighting of objective functions described in
Section~\ref{sec:objectiveweighting}. On my trip I had let {\tt
  playfun} run with particularly high constants ({\tt NFUTURES} = 50,
lots of backtracking candidates, etc.) and it had spent thousands of
CPU hours playing to this point, grabbing the coins, and then dying of
timeout, then losing all its lives, starting over, and getting to this
point again! After fixing the bug, I tried again, and it was a huge
breakthrough: Mario jumped up to get all the coins, and then almost
immediately jumped back down to continue the level! On his first try
he beat {\tt 1-2} and then immediately jumped in the pit at the
beginning of {\tt 1-3} just as I was starting to feel suuuuuper smart
(Figure~\ref{fig:mario13}). On a scale of OMFG to WTFLOL I was like
{\em whaaaaaaat}?

\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.95 \linewidth]{mario13}
\end{center}\vspace{-0.1in}
\caption{After making me feel so successful by finally cleanly
  finishing world {\tt 1-2}, Mario immediately jumps into the first
  trivial pit in {\tt 1-3}. I believe what's going on here is that
  this is actually the best future, because resetting back a single
  screen isn't a large loss, lives aren't included in the objective
  function, and probably the rest of the futures were struggling to
  make progress. This level is very jumpy and probably needs more
  futures to navigate its obstacles. In any case, Mario, Touch\'e.}
\label{fig:mario13}
\end{figure}

@@

{\bf Software engineering lesson:} Initial results from caching were
disappointing and I thought the overhead was to blame. I made several
low-level tweaks to avoid copying, etc., to reduce the overhead from
36\% to 13\%. Then I discovered that there were 0 cache hits ever,
because of a bug (I was inadvertantly using pointer equality on keys
instead of value equality, so keys were never found). Check basic
things about your code's correctness before going on an optimization
soiree!

@@

Before I had the idea that is the actual subject of this paper, I was
playing around with other emulator search ideas that involved storing
very very large numbers of states. This necessitated minimal
representations of savestates. FCEUX uses zlib internally to compress
states, which works well; they shrink from something like 13,776
bytes\footnote{I don't completely understand the NES architecture and
  emulation strategy, but I believe some games include expansion chips
  that require more space for save states. All this work is based on
  the 2048 bytes of main NES RAM, as you already know. Perhaps clever
  video game authors from the past can protect against this paper's
  approach by storing important game facts inside expansion chips
  in the cartridge.} to an average of 2266. I made some modifications
to the save/load routines to make this as small as I could manage.
I removed the backing buffer, which is only used for drawing graphics,
movie info, header bytes that are constant or already known (savestate
size), which shrunk the compressed size to an average of 2047.42 bytes.
That meant I could fit about 32 million savestates in RAM at once, which
is kind of amazing.

Compression algoritms work well when there is lots of redundancy in
their input. Comparing memories between two frames in some game,
you'll often see only a few differences. Some of the memory contains
stuff that's essentially read-only, even. To improve the
compressibility of savestates, I introduce the idea of a {\em basis}
state. This is any non-empty sequence of bytes which is an additional
argument to the {\tt LoadState} and {\tt SaveState} functions. It's
subtracted from the savestate prior to compression (if it's longer or
shorter, it's treated as an infinite repetition of those bytes). If
the basis is the same as the savestate, then the result is all zeroes,
which compresses nicely (of course, now you need to have the basis in
order to load the state, which is the same as the savestate, so that
didn't save you much!). But memories during gameplay often don't
differ much, so I save a state from any arbitrary frame in the game
and then use that as the basis. This factors out any common parts
and makes the states much more compressible: The average size drops
to 1870.41 bytes.

Using a better compression algorithm like the Burrows--Wheeler
transform\cite{Burrows}\footnote{This is really one of the all-time
  great algorithms. If you don't know it and you like this kind of
  thing, you should check it out. It's so surprising and elegant.}
would probably help too; zlip implements LZW which is only mediocre.
However, on a scale of throwing my computer out the window to wanting
to reimplement all software in my own private SML library, I just
can't stand getting 3\rd-party libraries to compile and link into my
applications, so I didn't even try. In any case, I abandoned this
direction, which was not working well even with lean savestates.

@@

% better to run on n-1 cpus?

@@

>>>>

\newcommand\comment[1]{}
\newcommand\parameteralert[1]{
  % XXX this can get cooler.
  \begin{quotation}
  \noindent {\bf Parameter Alert!} #1
  \end{quotation}
}

@@

The {\em improvability} of a state is the fraction of these potential
improvements that are legal improvements, as above. Based on my
analysis, states that are highly improvable ($> 30\%$) tend to be
``worse'' (more stuck, closer to death, or whatever) than those that
are not very improvable ($< 5\%$). This isn't being used for anything
currently, but getting an absolute picture of how good a state is (as
opposed to simply comparing it to an adjacent state) is one of the
difficulties with this approach, so this may be a useful notion for
future directions.

\paragraph{Takes of woe, tales of joy.} 
Backtracking was initially added in order to fix the mario coin ledge
problem in {\tt 1-2}. It did not help with this. Older versions of the
code would have Mario get past this spot, probably by luck, but in
these he would get stuck in an embarrassing way on a small pit later,
or jump straight into an enemy. Most of the approaches that actually
looked solidly good elsewhere were failing badly here. As I started
writing up this paper, on an airplane, I discovered the bug in the
weighting of objective functions described in
Section~\ref{sec:objectiveweighting}. On my trip I had let {\tt
  playfun} run with particularly high constants ({\tt NFUTURES} = 50,
lots of backtracking candidates, etc.) and it had spent about a
thousand CPU hours playing to this point, grabbing the coins, and then
dying of timeout, then losing all its lives, starting over, and
getting to this point again! After fixing the bug, I tried again, and
it was a huge breakthrough: Mario jumped up to get all the coins, and
then almost immediately jumped back down to continue the level! On his
first try he beat {\tt 1-2} and then immediately jumped in the pit at
the beginning of {\tt 1-3} just as I was starting to feel suuuuuper
smart (Figure~\ref{fig:mario13}). On a scale of OMFG to WTFLOL I was
like {\em whaaaaaaat}?

\begin{figure}[hb]
\begin{center}
\includegraphics[width=0.95 \linewidth]{mario13}
\end{center}\vspace{-0.1in}
\caption{After making me feel so successful by finally cleanly
  finishing world {\tt 1-2}, Mario immediately jumps into the first
  trivial pit in {\tt 1-3}. I believe what's going on here is that
  this is actually the best future, because resetting back a single
  screen isn't a large loss, lives aren't included in the objective
  function, and probably the rest of the futures were struggling to
  make progress. This level is very jumpy and probably needs more
  futures to navigate its obstacles. In any case, Mario, Touch\'e.}
\label{fig:mario13}
\end{figure}

@@

{\bf Software engineering lesson:} Initial results from caching were
disappointing and I thought the overhead was to blame. I made several
low-level tweaks to avoid copying, etc., to reduce the overhead from
36\% to 13\%. Then I discovered that there were 0 cache hits ever,
because of a bug (I was inadvertantly using pointer equality on keys
instead of value equality, so keys were never found). Check basic
things about your code's correctness before going on an optimization
sortie!

@@

Before I had the idea that is the actual subject of this paper, I was
playing around with other emulator search ideas that involved storing
very very large numbers of states. This necessitated minimal
representations of savestates. FCEUX uses zlib internally to compress
states, which works well; they shrink from something like 13,776
bytes\footnote{I don't completely understand the NES architecture and
  emulation strategy, but I believe some games include expansion chips
  that require more space for save states. All this work is based on
  the 2048 bytes of main NES RAM, as you already know. Perhaps clever
  video game authors from the past can protect against this paper's
  approach by storing important game facts inside expansion chips
  in the cartridge.} to an average of 2266. I made some modifications
to the save/load routines to make this as small as I could manage.
I removed the backing buffer, which is only used for drawing graphics,
movie info, header bytes that are constant or already known (savestate
size), which shrunk the compressed size to an average of 2047.42 bytes.
That meant I could fit about 32 million savestates in RAM at once, which
would be kind of amazing if I could tell my 8 year-old self that.

Compression algoritms work well when there is lots of redundancy in
their input. Comparing memories between two frames in some game,
you'll often see only a few differences. Some of the memory contains
stuff that's essentially read-only, even. To improve the
compressibility of savestates, I introduce the idea of a {\em basis}
state. This is any non-empty sequence of bytes which is an additional
argument to the {\tt LoadState} and {\tt SaveState} functions. It's
subtracted from the savestate prior to compression (it doesn't have to
be the same size; it's treated as an infinite repetition of those
bytes). If the basis is the same as the savestate, then the result is
all zeroes, which compresses nicely (of course, now you need to have
the basis in order to load the state, which is the same as the
savestate, so that didn't save you much!). But memories during
gameplay often don't differ much, so I save a state from any arbitrary
frame in the game and then use that as the basis for everything else.
This factors out any common parts and makes the states much more
compressible: The average size drops to 1870.41 bytes.

Using a better compression algorithm like the Burrows--Wheeler
transform\cite{Burrows}\footnote{This is really one of the all-time
  great algorithms. If you don't know it and you like this kind of
  thing, you should check it out. It's so surprising and elegant.}
would probably help too; zlib implements LZW which is only mediocre.
However, on a scale of throwing my computer out the window to wanting
to reimplement all software in my own private SML library, I just
can't stand getting 3\rd-party libraries to compile and link into my
applications, so I didn't even try. In any case, I abandoned this
direction, which was not working well even with lean savestates.

@@

The processes communicate over TCP/IP. I only run it on my local
computer, but it would be reasonable to run it on a cluster or
possibly even the Internet. I used SDL's SDL\_Net as a portable
network interface in the hopes of keeping it platform-agnostic (FCEUX
is basically portable so it would be a shame to make this
Windows-specific, though I am not eager to try to get this thing to
compile on other platforms, I gotta be honest). SDL is a nightmare to
get working with mingw in a 64-bit compile, as usual, and SDL\_Net
contained some bugs\footnote{In particular, the {\tt
    SDLNet\_TCP\_Recv} call is supposed to block until it has received
  the entire requested length, but it occasionally returns early.}
that I had to spend some time working around. Anyway, I'm just
complaining. For serializing requests and responses to bytes, I used
Google's Protocol Buffer library\cite{ProtoBuf}, which is designed for
this kind of thing.

Helpers all start up with the same ROM and motif and objectives
loaded, so that they can simulate whatever the master would
compute.\footnote{Only the master reweights motifs, since it needs a
  single sample of memories that we've observed. That means that in
  cases where a helper generates random motifs, it does so with
  respect to the original human weighting. There are some other small
  things I simply avoided doing because they would require too much
  coordination between processes or make the RPCs too large; MARIONET
  was part of the system for much of its development.} They just wait
on a port for the master to send a request, which can be either ``try
this set of {\tt nexts} with these futures and send me the results'' or
``try to find $n$ improvements of these inputs {\tt improveme} using
so-and-so approach.'' In either case we send the base state as a
serialized savestate.

The master does the outer loop of the algorithm and all the writing to
disk and so on. When it needs to do an expensive step like the inner
loop, it prepares a set of jobs and sends them to workers using a little
fork-join library ({\tt netutil.*}). The library keeps track of the
outstanding jobs and which helpers are working, and feeds them work
when they are idle, until all the jobs are complete. It's even got a
color ANSI progress bar and stuff like that.

Utilization is very good (Figure~\ref{fig:utilization}), and we get
almost a linear speedup with the number of processes, up to the number
of hardware threads the system supports (12). Actually, in addition
to the computer remaining usable for other stuff, $n-1$ helpers seems
to work a bit better than $n$, because each of the helpers is then able
to avoid preemption. Since the driver loop is synchronous, having one
laggard process slows the whole darn thing down.

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.75 \linewidth]{utilization}
\end{center}\vspace{-0.1in}
\caption{Utilization with 12 helpers and one master on a 6-core (12
  hyperthreads) processor. Looks good. Keeps the bedroom warm.}
\label{fig:utilization}
\end{figure}

MARIONET is a huge improvement and was what enabled using 40+ futures
in a reasonable amount of time, which is key to high-quality output. It's
obviously the way you want to run the software, if you can get it to
compile. There is one downside, though, which is that it reduces our
ability to get lucky cache hits across {\tt nexts} that are similar (or
have the same effect), and when replaying the winner as we commit to it.
It's worth it, but smartly deciding which helper gets which tasks (because
they share prefixes, for example) would save time. Running it all in one
process with a shared memory cache would be the best, as long as the
lock contention could be kept under control.

@@

\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.95 \linewidth]{tetris-tower}
\end{center}\vspace{-0.1in}
\caption{Would you hire this construction company? Death is imminent,
so {\tt playfun} pauses the game shortly after this and then doesn't
unpause it.}
\label{fig:tetris-tower}
\end{figure}

Tetris is a block dropping game, also known to the ancients. The
Nintendo implementation is known for being inferior to the unlicensed
Tengen version, which Nintendo tried to sue out of existence. Here I
try to automate the Nintendo version as a tribute to litigation.
Unsurprisingly, {\tt playfun} is awful at this game. Tetris has
several screens of menus, where the player can select between
different modes and theme musics and stuff like that. As an amusing
prelude to this disaster in tetromino planning, {\tt playfun} keeps
entering the menu and exiting back to the title screen rapidly, taking
several seconds to even start the game. (Like in Bubble Bobble, this
means that the start button is among the motifs.) Although the piece
dropping looks more or less natural (but it's hard to not be, as the
game drops the pieces for you), the placement is idiotic. This may be
because placing a piece gives you a small amount of points, which
probably looks like progress (Figure~\ref{fig:tetris-objectives}). As
the screen fills, there's a tiny amount of tetris-like gameplay,
probably as the end of the game becomes visible in some of the
futures. The end result is awful and {\tt playfun} makes zero lines
and certainly no Tetrises (Figure~\ref{fig:tetris-tower}). The only
cleverness is pausing the game right before the next piece causes the
game to be over, and leaving it paused. Truly, the only winning move
is not to play.
